Fourier-domain modulations and delays of gravitational-wave signals

### 2017-09-05 docs/paper_method.tex
Added option for simplified likelihood. Edits to the paper and updated plots.
The simplified likelihood is for frozen LISA, in the low-f approximation. The response becomes purely external for fixed masses. At setup the appropriate prefactor is computed once and for all. The likelihood computation becomes a very fast evaluation of a closed-form analytical expression.

\documentclass[aps,showpacs,%twocolumn,
prd,superscriptaddress,nofootinbib]{revtex4}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{bm}
\usepackage{color}
\usepackage{enumerate}
\usepackage{ulem}

@@

\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand\ud{{\mathrm{d}}}
\newcommand\uD{{\mathrm{D}}}
\newcommand\calO{{\mathcal{O}}}
\newcommand\calM{{\mathcal{M}}}
\newcommand\calF{{\mathcal{F}}}
\newcommand\calT{{\mathcal{T}}}
\newcommand\bfx{\mathbf{x}}
\newcommand{\ov}[1]{\overline{#1}}
\newcommand{\ph}[1]{\phantom{#1}}
\newcommand{\cte}{\mathrm{cte}}
\newcommand{\nn}{\nonumber}
\newcommand{\hatk}{\hat{k}}
\newcommand{\Hz}{\,\mathrm{Hz}}
\newcommand{\sinc}{\,\mathrm{sinc}}
\newcommand{\Msol}{M_{\odot}}
\newcommand{\tf}{t_{f}}
\newcommand{\Tf}{T_{f}}
\newcommand{\tfd}{t_{f}^{d}}

@@

With the recent detections of gravitational-wave signals emitted by the coalescences of binary black holes, announced by the LVC collaboration after the first obervation run of the two LIGO instruments~\cite{LIGO-theevent-2016,LIGO-christmasevent-2016,LIGO-O1BBH-2016}, gravitational-wave astronomy has entered its observational era. With the recent start of LIGO's second observation run, and later the extension of the ground-based detectors network with Virgo~\cite{Virgo}, KAGRA~\cite{KAGRA} and LIGO-India~\cite{INDIGO}, many more detections from compact object coalescences are to come, with binary black holes being expected to be the dominant sources~\cite{}. Moreover, the European Space Agency has issued a mission call for its L3 space mission with a tentative launch around 2034, adressing the science theme ``The Gravitational Universe'', possibly opening the way for a space-borne observatory based on the concept of the Laser Interferometer Space Antenna (LISA)~\cite{LISA09,LISA17}. Among other sources of gravitational waves, LISA will be able to detect and characterize comparable-mass binary black hole coalescences to cosmological distances over a wide range of masses, from high-redshift observations of supermassive black hole binaries with $M\sim 10^{7} \Msol$ down to the observation of LIGO-type sources with $M\sim 10^{2} \Msol$~\cite{Sesana16}.

@@

The plan of the paper is as follows. In Sec.~\ref{sec:motivation}, we will introduce our notations and conventions for Fourier-domain waveforms, motivate the problem of Fourier-domain modulation and introduce the relevant timescales for both the response of LISA-type detectors and the modulation of precessing signals. In Sec.~\ref{sec:formalism}, we present our general formalism, give its leading order approximation as well as higher-order corrections, and introduce new timescales based on the Fourier-domain signal. In Sec.~\ref{sec:quadphasedelay}, we refine the previous results for both the quadratic-in-phase corrections and the treatment of the delays, and make the connection to previous approaches of the problem. We then apply our formalism to the response of the LISA detector in Sec.~\ref{sec:LISA}, and to the case of signals from precessing binaries in Sec~\ref{sec:precession}. We discuss and summarize our results in Sec.~\ref{sec:discussion}.

@@

A useful representation of the gravitational waveform is given by its decomposition in spin-weighted spherical harmonics. The waveform emitted in the direction $(\Theta, \Phi)$, with its two polarizations $h_{+},h_{\times}$, can be decomposed as a superposition of modes as~\cite{Thorne80}
\be\label{eq:defmodes}
	h_{+} - i h_{\times} = \sum\limits_{\ell \geq 2} \sum\limits_{m=-\ell}^{\ell} {}_{-2}Y_{\ell m}(\Theta,\Phi) h_{\ell m} \,,
\ee
where the ${}_{-2}Y_{\ell m}(\Theta,\Phi)$ are spin-weighted spherical harmonics. For a non-precessing system, with a fixed orbital plane, the individual modes have the additional symmetry property $h_{\ell -m} = (-1)^{\ell} h_{\ell m}^{*}$. Since we will work in the Fourier domain, it will be useful to write the contributions of the individual modes to the Fourier transforms of the polarizations as
\begin{subequations}\label{eq:hpcfrommodes}
\begin{align}
	\tilde{h}_{+}(f) &= \frac{1}{2} \sum\limits_{\ell \geq 2} \sum\limits_{m=-\ell}^{\ell} \left( {}_{-2}Y_{\ell m} \widetilde{h_{\ell m}}(f) + {}_{-2}Y_{\ell m}^{*} \widetilde{h_{\ell m}}(-f)^{*} \right) \,, \\
	\tilde{h}_{\times}(f) &= \frac{i}{2} \sum\limits_{\ell \geq 2} \sum\limits_{m=-\ell}^{\ell} \left( {}_{-2}Y_{\ell m} \widetilde{h_{\ell m}}(f) - {}_{-2}Y_{\ell m}^{*} \widetilde{h_{\ell m}}(-f)^{*} \right) \,,
\end{align}
\end{subequations}
where we used $\widetilde{h_{\ell m}^{*}}(f) = \widetilde{h_{\ell m}}(-f)^{*}$. An additional approximation often used in waveform models consists in considering that the Fourier transforms $\tilde{h}_{\ell m}$ have support only on one side of the spectrum, either for positive of for negative frequencies depending on the sign of $m$. This holds in particular within the stationary phase approximation, as explained below. With our sign convention~\eqref{eq:defFT}, this approximation reads
\begin{align}\label{eq:approxnegativef}
	\tilde{h}_{\ell m} (f) &\simeq 0 \text{ for } f<0, \; m>0 \nn\,,\\
	\tilde{h}_{\ell m} (f) &\simeq 0 \text{ for } f>0, \; m<0 \,.
\end{align}

@@

\begin{figure}
  \centering
  \includegraphics[width=.48\linewidth]{plots/Af.pdf}
  \hspace{0.2cm}
  \includegraphics[width=.48\linewidth]{plots/Psif.pdf}
  \caption{Fourier-domain amplitude and phase in geometric units for an equal-mass, non-spinning system. The vertical line represents $\omega_{22}^{\rm peak}/2\pi$, the frequency corresponding to the time-domain frequency of the waveform at the peak amplitude. The starting frequency $Mf_{min} = 2.6\times 10^{-4}$ is appropriate for a 2-years observation of a $M=10^{6} M_{\odot}$ system by LISA, or for an observation by LIGO/Virgo of a $M=5.4 M_{\odot}$ system entering the instrument band at $10\Hz$. [add residuals of interpolation on $~300$ points, to illustrate the compression of Amp/Phase FD waveforms ? or show different spins/mass ratios ?]}
  \label{fig:ampphase}
\end{figure}

We recall that, with our sign convention, the effect of a positive shift in time $h_{\Delta t} (t) = h(t- \Delta t)$ is
\be
	\tilde{h}_{\Delta t}(f) = e^{2i\pi f \Delta t} \tilde{h}(f) \,.
\ee
If we also allow for a global shift in phase of the signal, which corresponds to a rotation of the unit vector pointing towards the observer in the frame of the gravitational waves source, a constant term can also be added (note that the effect of such a rotation gives different phase changes $e^{\pm i m \delta \phi}$ for the different modes of the radiation). We see that both a constant and a linear term in the Fourier-domain phase can be adjusted by changing extrinsic parameters. Thus, as is well known, apart from the relative phases of the modes for waveforms that include higher harmonics, the intrinsic information on the phasing is contained in the second derivative of the Fourier-domain phase $\Psi(f)$.

@@

We recall here an approximation widely used to model gravitational waves signals emitted by inspiralling binaries, the stationary phase approximation (thereafter SPA). It applies in general to chirping signals, and we refer the reader to~\ref{} for details. One first writes the time-domain signal in amplitude and phase form as $h(t) = a(t) e^{-2i\varphi(t)}$, where for gravitational wave signals $\varphi$ will correspond to the orbital phase, with the orbital frequency being $\omega = \dot{\varphi}$. In order to keep close to the notations used in the gravitational wave literature, we introduced a factor of 2 in the phase of the wave, which is appropriate for the dominant 22 harmonic of the signal. The approximation then applies to signals verifying the conditions
\be\label{eq:conditionSPA}
	\left| \frac{\dot{a}/a}{\omega} \right| \ll 1\,, \quad \left|\frac{\dot{\omega}}{\omega^{2}} \right| \ll 1\,, \quad \left| \frac{(\dot{a}/a)^{2}}{\dot{\omega}} \right| \ll 1 \,.
\ee
Since the integral~\eqref{eq:defFT} defining the Fourier transform is rapidly oscillatory unless the term $2\pi f t$ cancels the evolution of $-2\varphi(t)$, its support is well centered around the point of stationary phase. This determines the time-to-frequency correspondence in the stationary phase approximation, and leads to the definition of this time as an implicit function of frequency by the relation
\be\label{eq:deftfSPA}
	\omega(t_{f}^{\rm SPA}) = \pi  f \,.
\ee
For a chirping signal of increasing phase, $\omega>0$ and $\dot{\omega}>0$, there is a unique point of stationary phase, located in the positive frequency range $f>0$. Using the conditions above, one can formally expand the signal around $t_{f}^{\rm SPA}$ to quadratic order in time, according to
\be
	\tilde{h}_{\rm SPA} (f) \simeq a(t_{f}^{\rm SPA}) \exp\left[2i\pi f t_{f}^{\rm SPA}-2i\varphi(t_{f}^{\rm SPA}) \right]  \int \ud t \, e^{-i \dot{\omega} (t_{f}^{\rm SPA}) (t-t_{f}^{\rm SPA})^{2}} \,.
\ee
Note that to be able to treat the amplitude as a constant in the integral above, we used the third condition in~\eqref{eq:conditionSPA}. The resulting complex Gaussian integral yields\footnote{Note that the expressions given here are valid for the $22$ mode of the waveform.}
\begin{subequations}
\begin{align}
	\tilde{h}_{\rm SPA}(f) &= A_{\rm SPA}(f) e^{-i\Psi_{\rm SPA}(f)} \,, \\
	A_{\rm SPA}(f) &= a(t_{f}^{\rm SPA}) \sqrt{\frac{\pi}{\dot{\omega}(t_{f}^{\rm SPA})}} \,, \\
	\Psi_{\rm SPA}(f) &= 2\varphi(t_{f}^{\rm SPA}) - 2\pi f t_{f}^{\rm SPA} + \frac{\pi}{4} \,.
\end{align}
\end{subequations}
Ref.~\cite{Droz+99} evaluated the first correction to this approximation, within the context of post-Newtonian signals, and found that it can be considered as a term of the fifth post-Newtonian order, beyond the accuracy level of our current best models~\cite{BlanchetLiving}.

To understand the separation of timescales in the problem, it will be useful to have at hand the leading-order scaling laws for an inspiral (at the Newtonian order in the PN language). Although inaccurate for the purpose of waveform modelling, these Newtonian estimates will give a good qualitative description of the order of magnitude of the relevant timescales in the inspiral. For a binary with masses $m_{1}, m_{2}$, we define the total mass $m=m_{1}+m_{2}$ and the symmetric mass ratio $\nu = m_{1}m_{2}/m^{2}$. Introducing a time of coalescence $t_{c}$, the relations between the orbital frequency and phase and the time to coalescence $t_{c} - t$ are then given by
\begin{subequations}
\begin{align}
	\omega(t) &= \left[ \frac{256\nu}{5c^{5}} (Gm)^{5/3} (t_{c}-t) \right]^{-3/8} \,, \\
	\varphi(t) &= -\left[ \frac{c^{3}}{G m \nu^{3/5}} (t_{c}-t) \right]^{5/8} \,.
\end{align}
\end{subequations}
For this Newtonian inspiral, applying the SPA gives for the $22$ mode $\tilde{h}(f) = A_{\rm N}(f)e^{-i\Psi_{\rm N}(f)}$ with
\begin{subequations}
\begin{align}
	A_{N}(f) &= \frac{G^{2}m^{2} \pi}{Dc^{5}} \sqrt{\frac{2\nu}{3}} v^{-7/2}\,,\\
	\Psi_{\rm SPA}^{\rm N}(f) &= 2\pi f t_{0} - \phi_{0} + \frac{3}{128\nu v^{5}} \,, 
\end{align}
\end{subequations}
where $v=(G m \pi f/c^{3})^{1/3}$, $D$ is the luminosity distance to the observer, and where $t_{0}, \phi_{0}$ are constants. It is also customary to rewrite the above relations in terms of the chirp mass $\calM \equiv m\nu^{3/5}$, which is the only mass combination arising in the phase at the leading PN order.

@@

The response of a detector of the LISA type to an incident gravitational wave can be written in two different, equivalent forms, in terms of phase or frequency measurements. Here we will work with the second representation, which will prove more convenient for our purposes. Moreover, various notations and conventions have been used in the literature to label the spacecrafts and describe their orbits. We refer the reader to~\cite{Vallisneri04} for a comparative account on these various conventions. In this work, we will use the conventions of~\cite{Vallisneri04}, which were also used in the Mock LISA Data Challenges (MLDC)~\cite{}.

@@

Written in terms of the fractional laser frequency shifts between two spacecrafts, the elementary response of the detector reads~\cite{EW75, RCP04, Vallisneri04}
\be\label{eq:yslr}
	y_{slr} \equiv \frac{\nu_{r} - \nu_{s}}{\nu} = \frac{1}{2} \frac{n_{l}^{i}n_{l}^{j}}{1 - \hat{k}\cdot n_{l}} \left[ h_{ij}^{\rm TT}(t - \hat{k}\cdot p_{s}) - h_{ij}^{\rm TT}(t - \hat{k}\cdot p_{r}) \right]
\ee
As detailed below in Sec.~\ref{}, we will use a rigid instantaneous model, approximating the geometry of the constellation by a moving equilateral triangle (neglecting the flexing of the arms induced by corrections in the orbits), and evaluating all geometric factors in~\eqref{eq:yslr} at a single time $t$ (neglecting pointing-ahead corrections).

The first- and second-generation TDI observables are then built as combinations of these basic building blocks, evaluated at delayed times. Since in our rigid approximation these delays will take a simple form in Fourier domain, we will focus in this paper on the $y_{slr}$ observables. We will give in Sec.~\ref{subsec:modelLISA} more details on the response of LISA-type detectors, on the TDI observables, and on the approximations that enter the derivation of the basic response~\eqref{eq:yslr} above.

The structure of~\eqref{eq:yslr} is such that one can build the full signal from individual contributions of the form
\be
	s(t) = F(t) h(t + d(t))
\ee
Here, $h(t)$ represents one of the individual modes building the full gravitational wave signal (see Eq.~\eqref{eq:defmodes}), $d(t)$ represents the time-varying delays of the form $-\hat{k}\cdot p_{A}(t)$, and $F(t)$ incorporates all the relevant geometric prefactors.

@@

It was suggested by a number of authors~\cite{} that the effect of the precession can be modeled, to a good approximation, by a time-dependent rotation of a non-precessing waveform. In this approach, one introduces a precessing frame following the evolution of the plane of the orbit, and approximates the waveform in this precessing frame by using a non-precessing model. Different prescriptions have been proposed for the construction of the precessing frame itself~\cite{}.

If $(\alpha, \beta, \gamma)$ are the Euler angles relating the precessing frame to the inertial frame in the $(z,y,z)$ convention, the modes in the inertial frame $h_{\ell m}^{\rm I}$ are then related to the modes in the precessing frame $h_{\ell m}^{\rm P}$ by~\cite{Goldberg+67}
\be\label{eq:wignerrot}
	h_{\ell m}^{\rm I} = \sum\limits_{m=-\ell}^{\ell} D^{\ell *}_{m'm} (\alpha,\beta,\gamma) h_{\ell m'}^{\rm P} \,,
\ee
where the $D^{\ell}_{mm'}$ are Wigner matrices~\cite{}. Notice that there is no mixing of the modes with different values of $\ell$.

If we make the assumption that the precessing-frame waveform $h^{\rm P}$ is approximated by a non-precessing model that provides us with a smooth Fourier-domain amplitude and phase, then the problem reduces to computing the Fourier transform of the signal
\be\label{eq:defmodulationprec}
	s(t) = F(t) h(t) \,,
\ee
where the modulation function $F(t)$ is given by a Wigner matrix and depends on time through the Euler angles $(\alpha, \beta, \gamma)(t)$. We will discuss more in Sec.~\ref{subsec:modelPrec} the limitations of the assumptions used to perform the decomposition above.

In Eq.~\eqref{eq:defmodulationprec} above, the modulation function $F$ evolves on the precessional timescale, which evolves throughout the inspiral. The separation between the precessional timescale and the orbital timescale will be crucial for our analysis. In the limit of low frequencies, we will see in Sec.~\ref{} below that, although these timescales become more and more separated, the decrease in the chirping rate gives raise to a corrective contribution that does not vanish in this limit. In the other limit, as the system gets close to merger, the separation of timescales becomes weaker, and we will explore in Sec.~\ref{} the application of our formalism to the merger and ringdown phase.

@@

\subsection{Leading order: the local approximation}
\label{subsec:LLP}

@@

We now extend our approach beyond the LLP approximation. We will give a formal derivation of the  general formula for the higher-order corrections in~\eqref{eq:FDkernel}, and then specialize it to the special cases of interest.

>>>>

\documentclass[aps,showpacs,twocolumn,
prd,superscriptaddress,nofootinbib]{revtex4-1}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{bm}
\usepackage{color}
\usepackage{enumerate}
\usepackage{ulem}
\usepackage{tabularx}

@@

\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
%\usepackage[colorlinks=true,
%            linkcolor=green,
%            urlcolor=blue,
%            citecolor=red]{hyperref}
\usepackage[colorlinks=true,
            linkcolor=YellowOrange,
            urlcolor=RoyalBlue,
            citecolor=RedViolet]{hyperref}

\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand\ud{{\mathrm{d}}}
\newcommand\uD{{\mathrm{D}}}
\newcommand\calO{{\mathcal{O}}}
\newcommand\calM{{\mathcal{M}}}
\newcommand\calF{{\mathcal{F}}}
\newcommand\calT{{\mathcal{T}}}
\newcommand\calD{{\mathcal{D}}}
\newcommand\calA{{\mathcal{A}}}
\newcommand\bfx{\mathbf{x}}
\newcommand{\ov}[1]{\overline{#1}}
\newcommand{\ph}[1]{\phantom{#1}}
\newcommand{\cte}{\mathrm{cte}}
\newcommand{\nn}{\nonumber}
\newcommand{\hatk}{\hat{k}}
\newcommand{\Hz}{\,\mathrm{Hz}}
\newcommand{\sinc}{\,\mathrm{sinc}}
\newcommand{\Msol}{M_{\odot}}
\newcommand{\tf}{t_{f}}
\newcommand{\Tf}{T_{f}}
\newcommand{\tfd}{t_{f}^{d}}

\newcolumntype{C}[1]{>{\centering\arraybackslash}p{#1}}
\newcolumntype{L}[1]{>{\raggedright\arraybackslash}p{#1}}

\newcommand{\SM}[1]{{\color{Red} #1}}

@@

With the recent detections of gravitational-wave signals emitted by the coalescences of binary black holes, announced by the LVC collaboration after the first obervation run of the two LIGO instruments~\cite{LIGO-theevent-2016,LIGO-christmasevent-2016,LIGO-O1BBH-2016}, gravitational-wave astronomy has entered its observational era. With the recent start of LIGO's second observation run, and later the extension of the ground-based detectors network with Virgo~\cite{Virgo}, KAGRA~\cite{KAGRA} and LIGO-India~\cite{INDIGO}, many more detections from compact object coalescences are to come, with binary black holes being expected to be the dominant sources~\cite{} \SM{[C]}. Moreover, the European Space Agency has issued a mission call for its L3 space mission with a tentative launch around 2034, adressing the science theme ``The Gravitational Universe'', possibly opening the way for a space-borne observatory based on the concept of the Laser Interferometer Space Antenna (LISA)~\cite{LISA09,LISA17}. Among other sources of gravitational waves, LISA will be able to detect and characterize comparable-mass binary black hole coalescences to cosmological distances over a wide range of masses, from high-redshift observations of supermassive black hole binaries with $M\sim 10^{7} \Msol$ down to the observation of LIGO-type sources with $M\sim 10^{2} \Msol$~\cite{Sesana16}.

@@

The plan of the paper is as follows. In Sec.~\ref{sec:motivation}, we will introduce our notations and conventions for Fourier-domain waveforms, motivate the problem of Fourier-domain modulation and introduce the relevant timescales for both the response of LISA-type detectors and the modulation of precessing signals. In Sec.~\ref{sec:formalism}, we present our general formalism, give its leading order approximation as well as higher-order corrections, introduce new timescales based on the Fourier-domain signal, refine the previous results for both the quadratic-in-phase corrections and the treatment of the delays, and make the connection to previous approaches of the problem. We then apply our formalism to the response of the LISA detector in Sec.~\ref{sec:LISA}, and to the case of signals from precessing binaries in Sec~\ref{sec:precession}. We discuss and summarize our results in Sec.~\ref{sec:discussion}.

@@

A useful representation of the gravitational waveform is given by its decomposition in spin-weighted spherical harmonics. The waveform emitted in the direction $(\Theta, \Phi)$, with its two polarizations $h_{+},h_{\times}$, can be decomposed as a superposition of modes as~\cite{Thorne80}
\be\label{eq:defmodes}
	h_{+} - i h_{\times} = \sum\limits_{\ell \geq 2} \sum\limits_{m=-\ell}^{\ell} {}_{-2}Y_{\ell m}(\Theta,\Phi) h_{\ell m} \,,
\ee
where the ${}_{-2}Y_{\ell m}(\Theta,\Phi)$ are spin-weighted spherical harmonics. For a non-precessing system, with a fixed orbital plane, the individual modes have the additional symmetry property
\be\label{eq:symmetryhlminusm}
	h_{\ell -m} = (-1)^{\ell} h_{\ell m}^{*} \,.
\ee
Since we will work in the Fourier domain, it will be useful to write the contributions of the individual modes to the Fourier transforms of the polarizations as
\begin{subequations}\label{eq:hpcfrommodes}
\begin{align}
	\tilde{h}_{+}(f) &= \frac{1}{2} \sum\limits_{\ell \geq 2} \sum\limits_{m=-\ell}^{\ell} \left( {}_{-2}Y_{\ell m} \tilde{h}_{\ell m}(f) + {}_{-2}Y_{\ell m}^{*} \tilde{h}_{\ell m}(-f)^{*} \right) \,, \\
	\tilde{h}_{\times}(f) &= \frac{i}{2} \sum\limits_{\ell \geq 2} \sum\limits_{m=-\ell}^{\ell} \left( {}_{-2}Y_{\ell m} \tilde{h}_{\ell m}(f) - {}_{-2}Y_{\ell m}^{*} \tilde{h}_{\ell m}(-f)^{*} \right) \,,
\end{align}
\end{subequations}
where we used $\widetilde{h_{\ell m}^{*}}(f) = \tilde{h}_{\ell m}(-f)^{*}$. An additional approximation often used in waveform models consists in considering that the Fourier transforms $\tilde{h}_{\ell m}$ have support only on one side of the spectrum, either for positive of for negative frequencies depending on the sign of $m$. This holds in particular within the stationary phase approximation, as explained below. With our sign convention~\eqref{eq:defFT}, this approximation reads
\begin{align}\label{eq:zeronegativef}
	\tilde{h}_{\ell m} (f) &\simeq 0 \text{ for } f<0, \; m>0 \nn\,,\\
	\tilde{h}_{\ell m} (f) &\simeq 0 \text{ for } f>0, \; m<0 \,.
\end{align}

@@

%\begin{figure*}
%  \centering
%  \includegraphics[width=.48\linewidth]{plots/Af.pdf}
%  \hspace{0.2cm}
%  \includegraphics[width=.48\linewidth]{plots/Psif.pdf}
%  \caption{Fourier-domain amplitude and phase in geometric units for an equal-mass, non-spinning system. The vertical line represents $\omega_{22}^{\rm peak}/2\pi$, the frequency corresponding to the time-domain frequency of the waveform at the peak amplitude. The starting frequency $Mf_{min} = 2.6\times 10^{-4}$ is appropriate for a 2-years observation of a $M=10^{6} M_{\odot}$ system by LISA, or for an observation by LIGO/Virgo of a $M=5.4 M_{\odot}$ system entering the instrument band at $10\Hz$. [add residuals of interpolation on $~300$ points, to illustrate the compression of Amp/Phase FD waveforms ? or show different spins/mass ratios ?] \SM{[Plot perhaps not necessary ?]}}
%  \label{fig:ampphase}
%\end{figure*}

We recall that, with our sign convention, the effect of a positive shift in time $h' (t) = h(t- \Delta t)$ is
\be
	\tilde{h}'(f) = e^{2i\pi f \Delta t} \tilde{h}(f) \,.
\ee
If we also allow for a global shift in phase of the signal, which corresponds to a rotation of the unit vector pointing towards the observer in the frame of the gravitational waves source, a constant term can also be added (note that the effect of such a rotation gives different phase changes $e^{\pm i m \delta \phi}$ for the different modes of the radiation). We see that both a constant and a linear term in the Fourier-domain phase can be adjusted by changing extrinsic parameters. Thus, as is well known, apart from the relative phases of the modes for waveforms that include higher harmonics, the intrinsic information on the phasing is contained in the second derivative of the Fourier-domain phase $\Psi(f)$.

@@

We recall here an approximation widely used to model gravitational waves signals emitted by inspiralling binaries, the stationary phase approximation (thereafter SPA). It applies in general to chirping signals, and we refer the reader to~\ref{} for details. One first writes the time-domain signal in amplitude and phase form as $h(t) = a(t) e^{-2i\varphi(t)}$, where for gravitational wave signals $\varphi$ will correspond to the orbital phase, with the orbital frequency being $\omega = \dot{\varphi}$. In order to keep close to the notations used in the gravitational wave literature, we introduced a factor of 2 in the phase of the wave, which is appropriate for the dominant 22 harmonic of the signal. The approximation then applies to signals verifying the conditions
\be\label{eq:conditionSPA}
	\left| \frac{\dot{a}/a}{\omega} \right| \ll 1\,, \quad \left|\frac{\dot{\omega}}{\omega^{2}} \right| \ll 1\,, \quad \left| \frac{(\dot{a}/a)^{2}}{\dot{\omega}} \right| \ll 1 \,.
\ee
Since the integral~\eqref{eq:defFT} defining the Fourier transform is rapidly oscillatory unless the term $2\pi f t$ cancels the evolution of $-2\varphi(t)$, its support is well centered around the point of stationary phase. This determines the time-to-frequency correspondence in the stationary phase approximation, and leads to the definition of this time as an implicit function of frequency by the relation
\be\label{eq:deftfSPA}
	\omega(t_{f}^{\rm SPA}) = \pi  f \,.
\ee
For a chirping signal of increasing phase, $\omega>0$ and $\dot{\omega}>0$, there is a unique point of stationary phase, located in the positive frequency range $f>0$. Using the conditions above, one can formally expand the signal around $t_{f}^{\rm SPA}$ to quadratic order in time, according to
\begin{align}
	\tilde{h}_{\rm SPA} (f) &\simeq a(t_{f}^{\rm SPA}) \exp\left[2i\pi f t_{f}^{\rm SPA}-2i\varphi(t_{f}^{\rm SPA}) \right] \nn\\
	& \qquad \cdot \int \ud t \, e^{-i \dot{\omega} (t_{f}^{\rm SPA}) (t-t_{f}^{\rm SPA})^{2}} \,.
\end{align}
Note that to be able to treat the amplitude as a constant in the integral above, we used the third condition in~\eqref{eq:conditionSPA}. The resulting complex Gaussian integral yields\footnote{Note that the expressions given here are valid for the $22$ mode of the waveform.}
\begin{subequations}
\begin{align}
	\tilde{h}_{\rm SPA}(f) &= A_{\rm SPA}(f) e^{-i\Psi_{\rm SPA}(f)} \,, \\
	A_{\rm SPA}(f) &= a(t_{f}^{\rm SPA}) \sqrt{\frac{\pi}{\dot{\omega}(t_{f}^{\rm SPA})}} \,, \\
	\Psi_{\rm SPA}(f) &= 2\varphi(t_{f}^{\rm SPA}) - 2\pi f t_{f}^{\rm SPA} + \frac{\pi}{4} \,.
\end{align}
\end{subequations}
Ref.~\cite{Droz+99} evaluated the first correction to this approximation, within the context of post-Newtonian signals, and found that it can be considered as a term of the fifth post-Newtonian order, beyond the accuracy level of our current best models~\cite{BlanchetLiving}.

To understand the separation of timescales in the problem, it will be useful to have at hand the leading-order scaling laws for an inspiral (at the Newtonian order in the PN language). Although inaccurate for the purpose of waveform modelling, these Newtonian estimates will give a good qualitative description of the order of magnitude of the relevant timescales in the inspiral. For a binary with masses $m_{1}, m_{2}$, we define the total mass $m=m_{1}+m_{2}$ and the symmetric mass ratio $\nu = m_{1}m_{2}/m^{2}$. Introducing a time of coalescence $t_{c}$, the relations between the orbital frequency and phase and the time to coalescence $t_{c} - t$ are then given by
\begin{subequations}
\begin{align}
	\omega(t) &= \left[ \frac{256\nu}{5c^{5}} (Gm)^{5/3} (t_{c}-t) \right]^{-3/8} \,, \\
	\varphi(t) &= -\left[ \frac{c^{3}}{G m \nu^{3/5}} (t_{c}-t) \right]^{5/8} \,.
\end{align}
\end{subequations}
For this Newtonian inspiral, applying the SPA gives for the $22$ mode $\tilde{h}(f) = A_{\rm N}(f)e^{-i\Psi_{\rm N}(f)}$ with
\begin{subequations}
\begin{align}
	A_{N}(f) &= \frac{G^{2}m^{2} \pi}{Dc^{5}} \sqrt{\frac{2\nu}{3}} v^{-7/2}\,,\\
	\Psi_{\rm SPA}^{\rm N}(f) &= \phi_{0} - 2\pi f t_{0} - \frac{3}{128\nu v^{5}} \,, 
\end{align}
\end{subequations}
\SM{[Check sign of $\Psi$ ? Should have $<0$ 2nd der.] }
where $v=(G m \pi f/c^{3})^{1/3}$, $D$ is the luminosity distance to the observer, and where $t_{0}, \phi_{0}$ are constants. It is also customary to rewrite the above relations in terms of the chirp mass $\calM \equiv m\nu^{3/5}$, which is the only mass combination arising in the phase at the leading PN order.

@@

The response of a detector of the LISA type to an incident gravitational wave can be written in two different, equivalent forms, in terms of phase or frequency measurements. Here we will work with the second representation, which will prove more convenient for our purposes. Moreover, various notations and conventions have been used in the literature to label the spacecrafts and describe their orbits. We refer the reader to~\cite{Vallisneri04} for a comparative account on these various conventions. In this work, we will use the conventions of~\cite{Vallisneri04}, which were also used in the Mock LISA Data Challenges (MLDC)~\cite{} \SM{[C]}.

@@

Written in terms of the fractional laser frequency shifts between two spacecrafts, the elementary response of the detector reads~\cite{EW75, RCP04, Vallisneri04}
\begin{align}\label{eq:yslr}
	y_{slr} &\equiv \frac{\nu_{r} - \nu_{s}}{\nu} \nn\\
	&= \frac{1}{2} \frac{n_{l}^{i}n_{l}^{j}}{1 - \hat{k}\cdot n_{l}} \left[ h_{ij}^{\rm TT}(t - \hat{k}\cdot p_{s}) - h_{ij}^{\rm TT}(t - \hat{k}\cdot p_{r}) \right] \,.
\end{align}
As detailed below in Sec.~\ref{}, we will use a rigid instantaneous model, approximating the geometry of the constellation by a moving equilateral triangle (neglecting the flexing of the arms induced by corrections in the orbits), and evaluating all geometric factors in~\eqref{eq:yslr} at a single time $t$ (neglecting pointing-ahead corrections).

The first- and second-generation TDI observables are then built as combinations of these basic building blocks, evaluated at delayed times. Since in our rigid approximation these delays will take a simple form in Fourier domain, we will focus in this paper on the $y_{slr}$ observables. We will give in Sec.~\ref{subsec:modelLISA} more details on the response of LISA-type detectors, on the TDI observables, and on the approximations that enter the derivation of the basic response~\eqref{eq:yslr} above. \SM{[explicit expressions for TDI]}

The structure of~\eqref{eq:yslr} is such that one can build the full signal from individual contributions of the form
\be
	s(t) = F(t) h(t + d(t)) \,.
\ee
Here, $h(t)$ represents one of the individual modes building the full gravitational wave signal (see Eq.~\eqref{eq:defmodes}), $d(t)$ represents the time-varying delays of the form $-\hat{k}\cdot p_{A}(t)$, and $F(t)$ incorporates all the relevant geometric prefactors.

@@

It was suggested by a number of authors~\cite{} \SM{[C]} that the effect of the precession can be modeled, to a good approximation, by a time-dependent rotation of a non-precessing waveform. In this approach, one introduces a precessing frame following the evolution of the plane of the orbit, and approximates the waveform in this precessing frame by using a non-precessing model. Different prescriptions have been proposed for the construction of the precessing frame itself~\cite{} \SM{[C]}.

If $(\alpha, \beta, \gamma)$ are the Euler angles relating the precessing frame to the inertial frame in the $(z,y,z)$ convention, the modes in the inertial frame $h_{\ell m}^{\rm I}$ are then related to the modes in the precessing frame $h_{\ell m}^{\rm P}$ by~\cite{Goldberg+67}
\be\label{eq:wignerrotintro}
	h_{\ell m}^{\rm I} = \sum\limits_{m=-\ell}^{\ell} \calD^{\ell *}_{mm'} (\alpha,\beta,\gamma) h_{\ell m'}^{\rm P} \,,
\ee
where the $\calD^{\ell}_{mm'}$ are Wigner matrices~\cite{} \SM{[C]}. Notice that there is no mixing of the modes with different values of $\ell$.

If we make the assumption that the precessing-frame waveform $h^{\rm P}$ is approximated by a non-precessing model that provides us with a smooth Fourier-domain amplitude and phase, then the problem reduces to computing the Fourier transform of the signal
\be\label{eq:defmodulationprec}
	s(t) = F(t) h(t) \,,
\ee
where the modulation function $F(t)$ is given by a Wigner matrix and depends on time through the Euler angles $(\alpha, \beta, \gamma)(t)$. We will discuss more in Sec.~\ref{subsec:modelPrec} \SM{[R]} the limitations of the assumptions used to perform the decomposition above.

In Eq.~\eqref{eq:defmodulationprec} above, the modulation function $F$ evolves on the precessional timescale, which evolves throughout the inspiral. The separation between the precessional timescale and the orbital timescale will be crucial for our analysis. In the limit of low frequencies, we will see in Sec.~\ref{} \SM{[R]} below that, although these timescales become more and more separated, the decrease in the chirping rate gives raise to a corrective contribution that does not vanish in this limit. In the other limit, as the system gets close to merger, the separation of timescales becomes weaker, and we will explore in Sec.~\ref{} \SM{[R]} the application of our formalism to the merger and ringdown phase.

@@

\subsection{Leading order: the local approximation \SM{[Find a better name for this approximation ?]}}
\label{subsec:LLP}

@@

We now extend our approach beyond the local approximation. We will give a formal derivation of the  general formula for the higher-order corrections in~\eqref{eq:FDkernel}, and then specialize it to the special cases of interest.
